Prey movement shapes the development of predator expertise in a virtual bi-trophic system

### 2024-07-09 manuscript/manuscript.Rmd
intro up to date.

The capacity of predators to match their hunting tactic to their prey is expected to drive the outcome of predator-prey interactions. To do so, predators need to practice their tactics extensively to develop the expertise to be successful hunters. Yet, there are limited empirical assessments showing links between experience and hunting success at the individual and population levels due to the challenges of monitoring direct interactions in the wild. Here, we use a virtual predator-prey system (the game *Dead by Daylight*) to investigate how experience shapes individual and population hunting success in predators across repeated interactions with their prey. We show that predators optimized prey consumption as they gained experience. Prey speed was an important mediator of this relationship, resulting in differences among predators in the development of expertise. **Décrire le reste des résultats...**

Keywords: foraging behaviour, experience, learning, antipredator behaviour, virtual ecology, Dead by Daylight

@@

The ecological consequences of predation accrue to the sum of consumptive and nonconsumptive effects (CEs and NCEs) during predator-prey interactions [@preisser2007]. Recent advances in our understanding of the behavioural mechanisms (e.g. foraging mode, foraging specialization, reciprocal plasticity) driving (N)CEs at the individual and population level have allowed ecologists to predict their outcome as well as their effect on prey population dynamics [@wirsing.etal2021]. For example, empirical studies show that prey tend to increase vigilance at the expense of foraging in the presence of ambush predators (NCEs) [@preisser2007]. Because ambush predators are opportunistic hunters, they tend to consume a greater range of prey types (CEs) compared to cursorial predators [@glaudas2019].

There are, however, internal mechanisms shaping predator foraging decision-making that are often overlooked when studying (N)CEs in predator-prey interactions [@wooster.etal2023]. For instance, predators need to acquire expertise by practicing and learning the proper skills to successfully locate, select, and capture their prey [@dukas2019;@wooster.etal2023]. The aquisition of expertise is predicted to be paramount in systems where predators learn different hunting tactics (e.g. orca whales) [@heithaus2018]. Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by learning associative images or using cues from their prey and their environment [@edwards.jackson1994;@morse2000;@macdonald2007;@reid.etal2010;@wilson-rankin2015]. Thus, predators can integrate this information to increase their chances of locating prey or to use the tactic most suited for the type of prey encountered [refs].

Conversely, prey can use antipredator tactics that are effective to evade predators, such as rapid or unpredictable movements [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017;@szopa-comley.ioannou2022]. This results in reciprocal interactions where predators need to learn to predict the movement of their prey to successfully capture them. Yet, it may be impossible for a predator to learn unpredictable prey escape trajectories such that these antipredator tactics may interfere with a predator's acquisition of hunting expertise [troscianko.etal2018]. Unfortunately, there is only scarce evidence showing links between experience, prey antipredator behaviour, and prey consumption in natural predators, impeding our ability to accurately predict the consequences of predation on prey populations.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Similar to agent-based simulations, online videogames can be useful systems to mitigate these challenges and test specific ecological hypotheses. They provide controlled virtual environments to test ecological hypotheses [see @lymbery.etal2023] with the advantage of having real players that interact in the virtual space. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. The prey can increase their chances of survival by cooperating and moving fast to escape the predator [@cere.etal2021;@fraserfranco.etal2022;@santostefano.etal2024]. The game also ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@beauchamp2020]. These observations outline how ecological patterns can emerge from interactions in virtual systems with fixed rules. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we assess how repeated encounters with prey shapes predator hunting success using data from players in *Dead by Daylight*. First, we investigate how individual hunting success changes with repeated encounters (i.e. how individuals develop their hunting expertise). At the population level, we hypothesize that the predators' success will increase with experience up to a certain level where it will stabilize [@dukas2019]. However, we expect this pattern to vary among individuals due to differences in the movement of prey encountered. Therefore, we then investigate how prey movement influences the development of expertise. If predators encounter groups of prey with similar speeds, we predict that the gain in expertise will be similar among individuals. Alternatively, if predators encounter groups of prey with varying speeds, then the acquisition of expertise will vary among individuals.

@@

Before the start of a match, players (predator or prey) can choose an avatar with unique abilities that encourage specific play styles (e.g. bold vs cautious prey, or ambush vs roaming predator). During our study period, the game offered `r toString(format(length(unique(dat$avatar_id))))` predator avatars. The virtual environments are composed of fixed and procedurally generated habitat components, such as vegetation, mazes, and buildings. Some of these environments are larger than others, with varying structural complexity. However, predators display only minimal changes in behaviour and hunting success across the environments, probably due to a game feature enabling them to have visual cues on the resources [@fraserfranco.etal2022]. There were `r toString(format(length(unique(dat$environment_id))))` virtual game environments available for play during the study period. Details on the basic characteristics of predator avatars are available at https://deadbydaylight.fandom.com/wiki/Killers. Details on the size and structure of the different virtual environments are available at https://dbdmaps.com/ and https://deadbydaylight.fandom.com/wiki/Realms.

@@

#fit1 <- readRDS(here::here("outputs", "01_outputs_models", "A2_GAMM-rank.rds"))
#fit2 <- readRDS(here::here("outputs", "01_outputs_models", "A2_GAMM-speed-rank.rds"))
#
#fit1
#fit2
```

@@

>>>>

**revoir abstract pour être moins centré sur la tactique**

**revoir la phrase sur ce qu'on fait dans l'étude. Intégrer l'aspect des proies**

The acquisition of expertise is crucial for predators to be successful hunters. To do so, predators need to hone their skills and acquire knowledge through repeated and extensive practice. Yet, there are limited empirical assessments showing how predators acquire expertise through repeated encounters with their prey due to the challenges of monitoring direct interactions in the wild. Here, we use a virtual predator-prey system (the game *Dead by Daylight*) to investigate how experience shapes individual and population hunting success in human predators across repeated interactions with their prey. We show that predators optimized prey consumption as they gained experience, indicating that they acquired expertise with extensive practice. Prey speed was an important mediator of this relationship, driving differences among predators in the development of expertise. At the population-level, we found that faster prey could impair the acquisition of expertise by reducing hunting success. At the individual-level, our results show that prey speed shapes how individual predators acquire expertise. Therefore, our study outlines how prey antipredator behaviour can mediate the acquisition of expertise in predator populations.

Keywords: predator-prey, experience, learning, antipredator behaviour, virtual ecology, Dead by Daylight

@@

The ecological consequences of predation accrue to the sum of consumptive and nonconsumptive effects (CEs and NCEs) during predator-prey interactions [@preisser2007]. Recent advances in our understanding of the mechanisms (e.g. foraging mode, reciprocal plasticity, size-dependent foraging) driving (N)CEs at the individual and population level have allowed ecologists to predict their outcome as well as their effect on prey population dynamics [@Peckarsky.etal2008;@gravel.etal2013;@wirsing.etal2021]. For example, empirical studies show that prey tend to increase vigilance at the expense of foraging in the presence of ambush predators (NCEs) [@preisser2007], with such hunters consuming a greater range of prey types (CEs) compared to cursorial hunters [@glaudas2019].

There are, however, internal mechanisms contributing to predator foraging success that are often overlooked when studying (N)CEs in predator-prey interactions [@wooster.etal2023]. For instance, predators need to acquire expertise by practicing and learning the proper skills to successfully locate, select, and capture their prey [@dukas2019;@wooster.etal2023]. Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by learning associative images, or by using cues from their prey and their environment [@edwards.jackson1994;@morse2000;@macdonald2007;@reid.etal2010;@wilson-rankin2015]. Thus, predators can integrate this information to increase their chances of locating prey, or to use the attack tactic most suited for a given context.

On the other hand, prey use antipredator tactics such as rapid escapes to evade predators [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017]. This compels predators to develop swift motor skills to succesfully capture such prey. However, predators may be limited in their capacity to develop the necessary attributes (e.g. physical, physiological, neurological) for fast-paced hunting, which may impair the development of their hunting expertise. For example, experimental studies have shown that certain camouflage strategies, such as disruptive coloration, can impair expertise acquisition in humans and birds [@martin.etal2012;@troscianko.etal2013;@troscianko.etal2018]. Unfortunately, while our understanding of how predation risk influences information acquisition in prey has increased over the years [reviewed in @crane.etal2024], there is only scarce evidence showing links between prey antipredator strategies and the development of expertise in human and nonhuman predators. Therefore, our ability to accurately predict the (N)CEs of predation on prey populations remain limited.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Similar to agent-based simulations, online videogames can be useful systems to mitigate these challenges and test specific ecological hypotheses. They provide controlled virtual environments to test ecological hypotheses [see @lymbery.etal2023] with the advantage of having real players that interact in the virtual space. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. The prey can increase their chances of survival by cooperating and moving fast to escape the predator [@cere.etal2021;@fraserfranco.etal2022;@santostefano.etal2024]. The game also ellicits natural reactions in players such as freezing when predation is imminent (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@beauchamp2020]. These observations outline how fundamental ecological patterns can emerge from human interactions in virtual systems with fixed rules [@brosnan.etal2017;@kasumovic.etal2017]. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we assess how repeated encounters with prey shapes predator hunting success using data from players in *Dead by Daylight*. First, we investigate how individual hunting success changes with repeated encounters (i.e. how individuals develop their hunting expertise). At the population level, we hypothesize that the predators' success will increase with experience up to a certain level where it will stabilize [@dukas2019]. However, we expect this pattern to vary among individuals due to differences in the movement of prey encountered. Therefore, we then investigate how prey movement influences the development of expertise. If predators encounter groups of prey with similar speeds, we predict that the gain in expertise will be similar among individuals. Alternatively, if predators encounter groups of prey with varying speeds, then the acquisition of expertise will vary among individuals. We predict that faster prey will impair expertise acquisition, such that predators that encountered faster prey will have lower success throughout experience.

@@

Before the start of a match, players (predator or prey) can choose an avatar with unique abilities that encourage specific play styles (e.g. bold vs cautious prey, or ambush vs roaming predator). During our study period, the game offered `r toString(format(length(unique(dat$avatar_id))))` predator avatars. The virtual environments are composed of fixed and procedurally generated habitat components, such as vegetation, mazes, and buildings. Some of these environments are larger than others, with varying structural complexity. However, predators display only minimal changes in behaviour and hunting success across these environments, probably due to a game feature enabling them to have visual cues on the resources [@fraserfranco.etal2022]. There were `r toString(format(length(unique(dat$environment_id))))` virtual game environments available for play during the study period. Details on the basic characteristics of predator avatars are available at https://deadbydaylight.fandom.com/wiki/Killers. Details on the size and structure of the different virtual environments are available at https://dbdmaps.com/ and https://deadbydaylight.fandom.com/wiki/Realms.

@@

# Import top 2 model fits
fit1 <- readRDS(here::here("outputs", "outputs_models", "A2_GAMM-rank.rds"))
fit2 <- readRDS(here::here("outputs", "outputs_models", "A2_GAMM-speed-rank.rds"))
```

@@

```{r fig2, out.width = "100%", fig.cap = "**Figure 2**. Posterior distributions of the individual-level standard deviation parameters. The first two panels represent the intercept and slope standard deviations of the linear components relating hunting success to cumulative experience, and the third panel represents the standard deviation of the wiggliness component of the curves. The light coloured distributions are for the model with a shared trend where we did not account for prey speed, and the darker coloured distributions are for the model with a shared trend where we accounted for prey speed (i.e. the model with the highest predictive power)"}

knitr::include_graphics(
  here::here("outputs", "outputs_figures", "figure2.png")
)
```
